---
title: "601 Final Group Project"
author: "Tim, Noah, Caleb, Joseph"
date: "2024-12-06"
output: html_document
---


```{r load packages}
library(sf)
library(dplyr)
library(ggplot2)
library(leaflet)
```

#Set Working Directory
```{r Create Path}
#Load in SVI Data
SVI <- read.csv("Maryland.csv")

head(SVI)

#Load LIHTC Census Tract Data
BQT <- st_read(dsn = "Baltimore_City_Qualified_Census_Tracts.shp")
```



### 1. Handling Missing Values (if any)
```{r missing data}
#Check missing values for SVI Data
colSums(is.na(SVI))

#Check missing values for LIHTC Census Tract Data
colSums(is.na(BQT))

#Filter and view missing data
na.AH <- BQT %>%
  filter(if_any(everything(), is.na))

#Filter and remove -999 value in EP_HBURD
SVI <- SVI %>% filter(EP_HBURD != -999)
```


### 2. Transform and Filter Variables (if necessary)
```{r transform and filter}

#Histogram of percentage of persons (age 5+) who speak English “less than well”
hist(SVI$EP_LIMENG)

#Create column for square root of EP_LIMENG
SVI <- SVI %>%
      mutate(ROOT_EP_LIMENG = sqrt(EP_LIMENG))

#Histogram of root of percentage percentage of persons (age 5+) who speak English “less than well”
hist(SVI$ROOT_EP_LIMENG)


#Histogram of percentage minority
hist(SVI$EP_MINRTY)

# Create categorical column for greater than 60% minority (median) or less than 60% minority
SVI <- SVI %>%
        mutate(MEDIAN_MINRTY = ifelse(EP_MINRTY >60, TRUE, FALSE))

#Histogram of percentage minority by whether the tract has a minority population greater than the median
hist(SVI$EP_MINRTY[SVI$MEDIAN_MINRTY == TRUE])

hist(SVI$EP_MINRTY[SVI$MEDIAN_MINRTY == FALSE])

#Histogram of number of tracts with a minority population greater than the median
MINRTY_COUNT <- SVI %>%
                  group_by(MEDIAN_MINRTY) %>%
                  summarize(COUNT = n())

ggplot(MINRTY_COUNT, aes(x = MEDIAN_MINRTY, y = COUNT)) +
  geom_bar(stat = "identity")

#Histogram of count minority
hist(SVI$E_MINRTY)

#Create column for log of E_MINRTY
SVI <- SVI %>%
     mutate(LOG_E_MINRTY = log(E_MINRTY))

#Histogram of log of count minority
hist(SVI$LOG_E_MINRTY)


#Histogram of Housing Burdened Percentage
hist(SVI$EP_HBURD)


#Filter and select SVI data
SVI_FILTER <- SVI %>%
              #Filter for variables we are analyzing
              select(FIPS, LOCATION, AREA_SQMI, EP_LIMENG, ROOT_EP_LIMENG,
                     EP_MINRTY, MEDIAN_MINRTY, E_MINRTY, LOG_E_MINRTY, EP_HBURD)
```

## Step 4: Descriptive Statistics and Distribution Analysis

Descriptive stats give you a sneak preview of your data’s “personality,” and visualizing distributions helps you spot patterns or surprises.

### 1. Summary Statistics
``` {r summary stats}
SVI_LIM_MIN_GROUP_SUMMARY <- SVI_FILTER %>%
  select(EP_LIMENG, ROOT_EP_LIMENG, EP_MINRTY, MEDIAN_MINRTY, E_MINRTY, LOG_E_MINRTY) %>%
  summary()

SVI_LIM_MIN_GROUP_SUMMARY
```

```{r minority distributions}
#Distribution of Percentage Minority
ggplot(SVI_FILTER, aes(EP_MINRTY)) +
  geom_density(color = "darkgrey", fill = "grey") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(linetype = "solid")
  ) +
  scale_x_continuous(breaks = seq(0,100,10), limits = c(0,100)) +
  labs(title = "Distribution of Perecentage Minority",
       x = "Perecentage Minority",
       y = "Density") +
  geom_vline(aes(xintercept = mean(EP_MINRTY)), color = "blue", alpha = 0.7) +
  geom_vline(aes(xintercept = median(EP_MINRTY)), color = "darkorange", alpha = 0.7) +
  geom_text(aes(x=mean(EP_MINRTY), label="\nMean", y=0.012), colour="blue", angle=90) +
  geom_text(aes(x=median(EP_MINRTY), label="Median\n", y=0.012), colour="darkorange", angle=90)

#Plot number of tracts with majority minority populations
ggplot(MINRTY_COUNT, aes(x = MEDIAN_MINRTY, y = COUNT)) +
  geom_bar(stat = "identity", color = "steelblue", fill = "skyblue") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(linetype = "solid")
  ) +
  labs(title = "Count of Tracts with Percentage Minority Greater or Less Than Median Percentage (60%)",
       x = "Does the Tract Have a Percentage Minority Greater than the Median Percentage (60%)?",
       y = "Count")

#Distribution of Count Minority (log)
ggplot(SVI_FILTER, aes(LOG_E_MINRTY)) +
  geom_density(color = "darkgrey", fill = "grey") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(linetype = "solid")
  ) +
  scale_x_continuous(breaks = seq(5,10,0.5), limits = c(5,10)) +
  labs(title = "Distribution of Minority Population (log)",
       x = "Count of Minorities (log)",
       y = "Density") +
  geom_vline(aes(xintercept = mean(LOG_E_MINRTY)), color = "blue", alpha = 0.7) +
  geom_vline(aes(xintercept = median(LOG_E_MINRTY)), color = "darkorange", alpha = 0.7) +
  geom_text(aes(x=mean(LOG_E_MINRTY), label="\nMean", y=0.2), colour="blue", angle=90) +
  geom_text(aes(x=median(LOG_E_MINRTY), label="Median\n", y=0.2), colour="darkorange", angle=90)
```



### 3. Spatial Exploration with Maps

```{r choropleth}
#Percentage of persons (age 5+) who speak English “less than well” 
#Showing english profiency on a map of baltimore
EP_LIMENG_MAP <- ggplot() +
  geom_point(data = SVI, aes(fill = ROOT_EP_LIMENG), linewidth = 0, alpha = 0.9) +
  theme_void() +
  scale_fill_viridis_c(option = "B",
    name = "Perecentage of Housing\nCost-burdened Occupied\nHousing Units (root)",
    ) +
    labs(
    title = "Cost-burdened Housing Units Across Baltimore.",
  )

print = EP_LIMENG_MAP

#Percentage Minority
#Same but for minorities
MINRTY_MAP <- ggplot() +
  geom_sf(data = SVI, aes(fill = EP_MINRTY), linewidth = 0, alpha = 0.9) +
  theme_void() +
  scale_fill_viridis_c(option = "B",
    name = "Perecentage Minority",
    ) +
    labs(
    title = "Minority Percentage Across Balitmore",
  )


#Same but for minority count
MINRTY_C_MAP <- ggplot() +
  geom_sf(data = SVI, aes(fill = E_MINRTY), linewidth = 0, alpha = 0.9) +
  theme_void() +
  scale_fill_viridis_c(option = "B",
    name = "Minority Count",
    ) +
    labs(
    title = "Minority Count Across Baltimore",
  )


#Same but for minority median
MINRTY_MEDIAN_MAP <- ggplot() +
  geom_sf(data = SVI, aes(fill = MEDIAN_MINRTY), color = "grey", linewidth = 0.05, alpha = 0.9) +
  theme_void() +
  scale_fill_viridis_d(option = "D",
    name = "Perecentage Minority Greater\nThan Median Percentage (60%)",
    ) +
    labs(
    title = "Minority Population Across Balitmore.",
    subtitle = "Perecentage Minority by Median Percentage (60%)"
  )


```


## Step 5: Hypothesis Testing

### 1. T-tests for Continuous Data
```{r T-test hypothesis}
# T-test 1
# Create a categorical variable from MEDIAN_MINRTY (High/Low Minority)
median_minrty <- median(SVI$MEDIAN_MINRTY, na.rm = TRUE)

# Create a new column 'MINRTY_GROUP' based on whether MEDIAN_MINRTY is greater or less than the median
SVI <- SVI %>%
  mutate(MINRTY_GROUP = ifelse(MEDIAN_MINRTY >= median_minrty, "High Minority", "Low Minority"))

# Perform t-test comparing EP_LIMENG for "High Minority" vs "Low Minority"
t_test_minrty <- t.test(EP_LIMENG ~ MINRTY_GROUP, data = SVI)

# Print the t-test result for High vs Low Minority
print("T-test for High vs Low Minority:")
print(t_test_minrty)
```


### 3. One-Way ANOVA
```{r ANOVA One-Way}
# Possible: Housing Burden by zoning types
# Zoning types and EP Minority  
# Zoning types and EP English Prof   <- explore these in summary statistics/explore distributions too
```

## Step 6: Correlation Analysis

# 
```{r}
# CORRELATION MATRIX OF EP MINORITY, EP LIMENG, AND EP HBURD
cor_matrix <- SVI_FILTER %>%
              select(EP_MINRTY, ROOT_EP_LIMENG, EP_HBURD) %>%  
              cor(use = "complete.obs")  
print(cor_matrix)


# SCATTERPLOT CORRELATION BETWEEN EP LIMENG AND EP MINRTY
#Correlation Matrix between Language Proficiency and Minority
cor_1 <- SVI_FILTER %>%
         select(ROOT_EP_LIMENG, EP_MINRTY)
  
cor_matrix_1 <- cor(cor_1[, c("ROOT_EP_LIMENG", "EP_MINRTY")])
cor_matrix_1

#Correlation graph between Language Proficiency and Minority
ggplot(cor_1, aes(EP_MINRTY, ROOT_EP_LIMENG)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue", se = FALSE)


# SCATTERPLOT CORRELATION BETWEEN EP MINORITY AND EP HBURD
cor_2 <- SVI_FILTER %>%
         select(EP_MINRTY, EP_HBURD)
  
cor_matrix_2 <- cor(cor_2[, c("EP_MINRTY", "EP_HBURD")])
cor_matrix_2

#Correlation graph between Language Proficiency and Minority
ggplot(cor_2, aes(EP_MINRTY, EP_HBURD)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue", se = FALSE)


# SCATTERPLOT CORRELATION BETWEEN EP LIMENG AND EP HBURD
cor_3 <- SVI_FILTER %>%
         select(ROOT_EP_LIMENG, EP_HBURD)
  
cor_matrix_3 <- cor(cor_3[, c("ROOT_EP_LIMENG", "EP_HBURD")])
cor_matrix_3

#Correlation graph between Language Proficiency and Minority
ggplot(cor_3, aes(ROOT_EP_LIMENG, EP_HBURD)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue", se = FALSE)
```


## Multilinear Regression 
```{r}
# MULTILINEAR REGRESSION BETWEEN EP MINORITY, EP LIMENG, AND EP HBURD
# Fit the multilinear regression model
model <- lm(EP_HBURD ~ EP_MINRTY * ROOT_EP_LIMENG, data = SVI_FILTER)
summary(model) # Check model summary

# Visualize Both Variables vs. EP_HBURD in 1 graph (Not related to the rest of this section)
ggplot(SVI_FILTER, aes(x = EP_MINRTY, y = EP_HBURD, color = ROOT_EP_LIMENG)) +
  geom_point() +
  labs(
    title = "EP_MINRTY and ROOT_EP_LIMENG vs. EP_HBURD",
    x = "EP_MINRTY",
    y = "EP_HBURD"
  ) +
  stat_smooth(method="lm",se=FALSE)

# Add Predicted Values
SVI_FILTER$Predicted <- predict(model, newdata = SVI_FILTER)

# Visualize the Multiple Linear Regression
ggplot(SVI_FILTER, aes(x = EP_HBURD, y = Predicted)) +
  geom_point(color = "blue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Observed vs. Predicted Housing Burden % from Minority % and Lim Eng Prof. %",
    x = "Observed EP_HBURD",
    y = "Predicted EP_HbURD"
  ) +
  theme_minimal()

# Comprehensive diagnostic plots
par(mfrow = c(2, 2))
plot(model)
```

